Evaluating Hydrogen and Carbon Network Strategies for the European Energy System

### 2023-12-20 report/submissions/journal/paper/paper.tex
convert "models" into "scenarios"

% make abbreviation for co2
\newcommand{\COtwo}{CO$_2$}
\newcommand{\Htwo}{H$_2$}
\newcommand{\COgrid}{CO$_2$\=/Grid}
\newcommand{\Hgrid}{H$_2$\=/Grid}
\newcommand{\modBase}{Baseline model}
\newcommand{\modCO}{CO$_2$\=/Grid model}
\newcommand{\modH}{H$_2$\=/Grid model}
\newcommand{\modHybrid}{Hybrid model}

% alternative
\newcommand{\carbon}{CO$_2$}
\newcommand{\hydrogen}{H$_2$}
\newcommand{\carbongrid}{CO$_2$\=/Grid}
\newcommand{\hydrogengrid}{H$_2$\=/Grid}
\newcommand{\baselinemodel}{Baseline model}
\newcommand{\carbonmodel}{CO$_2$\=/Grid model}
\newcommand{\hydrogenmodel}{H$_2$\=/Grid model}
\newcommand{\hybridmodel}{Hybrid model}
\newcommand{\baselinescenario}{Baseline scenario}
\newcommand{\carbonscenario}{CO$_2$\=/Grid scenario}
\newcommand{\hydrogenscenario}{H$_2$\=/Grid scenario}
\newcommand{\hybridscenario}{Hybrid scenario}

@@

\title{Evaluating \Htwo{} and \COtwo{} Network Strategies for the European Energy System}
% \title{Competition and Synergies of \Htwo{} and \COtwo{} Networks in Europe}

@@

\begin{abstract}
    Hydrogen and carbon dioxide transport can both play an important role in climate-neutral energy systems. Hydrogen networks help serve regions with high energy demand, while excess emissions are transported away in carbon dioxide networks. When it comes to the synthesis of carbonaceous fuels, it is less clear which input should be transported: hydrogen to carbon point sources or carbon to low-cost hydrogen. We explore the potential synergies and competition of both networks in a cost-optimal carbon-neutral European energy system. In direct comparison, a hydrogen network is more cost-effective than a carbon network, since it serves to transport hydrogen to demand and to point source of carbon for utilization. However, in a hybrid scenario where both networks are present, the carbon network effectively complements the hydrogen network, promoting carbon capture from biomass and reducing reliance on direct air capture. This study shows how the European energy system in future net-zero and net-negative carbon dioxide emissions scenarios can benefits from integrating both networks.
\end{abstract}

@@

The transition to a climate-neutral European economy is a pressing challenge that demands coordinated action across various energy sectors. While management of both carbon dioxide (\COtwo{}) and hydrogen (\Htwo{}) is considered a critical component of this transition, a gap exists in understanding how new hydrogen infrastructure effectively interacts with carbon management technologies, including carbon capture, transport, storage, utilization, and sequestration. Hydrogen is consider in several sectors in industry not amenable to electrification, and offers an efficient way to transport and store energy over long distances and timeframes, and can be produced from electrolysis or steam methane reforming (SMR), for example. Captured carbon can be stored in geological formations to permanently extract it from the atmosphere, a process known as carbon sequestration (CS). Additionally, carbon can be combined with hydrogen to produce carbonaceous fuels to supply for example synthetic kerosene for aviation, synthetic methanol for shipping or synthetic methane for industrial feedstocks. From hereon, we jointly refer to the production of these carbonaceous fuels as Carbon Utilization (CU). Carbon can be effectively captured from industrial processes and the combustion of biomass, fossil fuels, or synthetic carbonaceous fuels through point-source carbon capture (CC) techniques, or harvested from the atmosphere using direct air capture (DAC). Both \COtwo{} and \Htwo{} networks are likely to play a crucial role in the cost-effective integration of these technologies to enable net-zero economies.

Recently, policymakers and industry in Europe have started developing hydrogen and carbon management strategies~\cite{GermanyDevelopingStrategy2023,CarbonManagementStrategie}, planning infrastructure components~\cite{CONetz}, and committing to the first carbon utilization projects~\cite{EFuelsPilotPlant2022,OrstedAssumesFull,GROUNDBREAKINGEFUELPRODUCTION,DLREfuelsDLR}. With the European Union's goal of achieving climate neutrality by 2050, a wide range of programs, funding models and initiatives have been established with dedicated instruments supporting both the ramp-up of the hydrogen and carbon economy~\cite{eu2023netzero,europeangreendeal,europeaninnovationfund}.
Collaborative industry initiatives like the European Hydrogen Backbone~\cite{gasforclimateEuropeanHydrogenBackbone2022} or the Hydrogen Infrastructure Map~\cite{H2InfrastructureMap} showcase the potential of hydrogen as a fuel and energy carrier. Some natural gas pipelines have already been repurposed to transport hydrogen~\cite{RohrFreiFuer}. Business models from companies like Tree Energy Solutions~\cite{TESHydrogenLife2023}, Carbfix~\cite{WeTurnCO2}, and Equinor~\cite{adomaitisEquinorRWEBuild2023} advertise carbon management hubs that provide green hydrogen, methane, and synthetic fuels on the one hand and offer \COtwo{} offtake on the other hand. The Northern Lights project in Norway~\cite{NorthernLightsWhat} is planning with a transport and sequestration capacity of 1.5~Mt \COtwo{} per year to be operative in 2024, expanding to a targeted scale of 5~Mt per year of sequestration by 2030.
The market potential for carbon capture and sequestration in Europe is underscored by the Capture Map's estimated potential~\cite{ToolsGreenTransition} of 1.7 Gt carbon annually from point sources, accounting for about half of the continent's emissions, as well as estimated sequestration capacity potentials of up to 3~Gt/a~\cite{europeancommissionEuropeanCO2Storage}. However, it's important to note that carbon sequestration technology, unlike carbon capture, is not yet fully mature and its full potential and implications remain to be explored in depth.
To advance the carbon economy, the Clean Air Task Force emphasizes the importance of developing a carbon transport system in Europe alongside a hydrogen network~\cite{lockwoodEuropeanStrategyCarbon}. Carbon pipelines, considered a mature technology, have seen widespread installations in the United States and Canada, primarily to supply \carbon{} to enhanced oil recovery~\cite{righettiSitingCarbonDioxide2017,friedmannNETZEROGEOSPHERICRETURN}.

Up to this point, it remains unclear how the two transport systems of hydrogen and carbon may complement or even compete with each other. Both can bridge the misalignment of sources and sinks for carbon and hydrogen. A hydrogen network can supply regions with geographically fixed hydrogen demand, such as for steelmaking, with hydrogen from regions with the best renewable resources as well as enable CU at the site of point-source CC. On the other hand, the carbon transport system allows for transporting captured carbon to regions with sequestration potentials or high-quality renewable resources, the latter enabling cost-effective CU.

In the literature, the two network approaches and underlying technologies have been discussed in several of publications, all of which, however, dealt with the isolated aspects~\cite{bakkenLinearModelsOptimization2008,morbeeOptimisedDeploymentEuropean2012,stewartFeasibilityEuropeanwideIntegrated2014,oeiModelingCarbonCapture2014,elahiMultiperiodLeastCost2014,burandtDecarbonizingChinaEnergy2019,middletonSimCCSOpensourceTool2020,bjerketvedtOptimalDesignCost2020,weiProposedGlobalLayout2021,damoreOptimalDesignEuropean2021,becattiniCarbonDioxideCapture2022,neumannBenefitsHydrogenNetwork2022}. Such techno-economic models, in contrast for example to Integrated Assessment Models, can account for the spatial distribution of carbon sources and sinks which are crucial to provide a holistic view of the energy system and its technological interactions. Neumann et al.~\cite{neumannBenefitsHydrogenNetwork2022} demonstrated the interchangeability of hydrogen and electricity grid expansion in a climate-neutral European energy system at the expense of higher investments. The underlying, highly-resolved model encompasses most relevant sectors, however it neglects the option to transport carbon across regions.
Morbee et al.~\cite{morbeeOptimisedDeploymentEuropean2012} optimizes the topology and capacity of a \COtwo{} network in Europe, but only considers the power sector without co-optimizing renewable deployment. This limited sectoral scope cannot capture important dynamics of carbon management, since it neglects the sectors like industry that will need to handle most \COtwo{} in the future.
A comprehensive example is found in~\cite{becattiniCarbonDioxideCapture2022}, which presents a mixed-integer model to optimize the time-evolution of a \COtwo{} transport system in Switzerland, connecting to a remote sequestration site in Norway. However, this limited spatial scope fails to consider other sequestration sites and co-benefits from connecting the \COtwo{} network to neighboring countries.
Other high-resolution energy system models do not feature detailed CU technologies and \COtwo{} transport. Euro-Calliope, for example, does neither feature carbon nor hydrogen transport, and thus forces generic synthetic fuel production to rely on captive electrolysis and DAC in the same region~\cite{pickeringDiversityOptionsEliminate2022}.

No study has yet considered the co-optimization and comprehensive assessment of both \COtwo{} and \Htwo{} networks in a fully sector-coupled energy system. However, such an assessment is strongly needed to identify realms of competition and synergy between hydrogen and carbon management technologies. In this paper, we present a detailed study of the European energy system for 2050, which includes high geographical resolution and a comprehensive representation of carbon management technologies. The study is conducted using the PyPSA-Eur energy system model which optimizes operation and investments in all relevant energy sectors to supply projected energy demands for 2050 (see Fig.~\ref{fig:total-demand-bar}). It is assumed that Europe is self-sufficient in energy and does not import any fuels. This drives carbonaceous fuels production to be located at sites where both \carbon{} and \hydrogen{} can be provided at low cost.

@@

To investigate the competing transport dynamics between \COtwo{} and \Htwo{} networks, we contrast the following four model scenarios:
%
\begin{itemize}
    \item[] \textit{Baseline:} Neither \COtwo{} nor \Htwo{} networks are constructed.
    \item[] \textit{\COgrid{}:} Only the \COtwo{} network is developed.
    \item[] \textit{\Hgrid{}:} Only the \Htwo{} network is developed.
    \item[] \textit{Hybrid:} Both \COtwo{} and \Htwo{} networks are developed.
\end{itemize}

@@

All scenarios share key energy system characteristics but differ in system costs and technology use (Fig.~\ref{fig:cost_bar}). The \modBase{} has the highest annual cost at 779~bn€/a. Models with \carbon{} and \hydrogen{} transport options are more cost-effective, reducing costs by 3.1\% in \modCO{}, 4.8\% in \modH{}, and 5.2\% in \modHybrid{}.

\begin{figure}[ht!]
    \centering
    \includegraphics[width=\linewidth]{comparison/default/figures/90_nodes/cost_bar.png}
    \caption[short]{Total annual system cost subdivided into groups of technologies for the different models of the European energy system with a net-zero emission target. While the \modBase{} has neither a \COtwo{} nor an \Htwo{} network, the \modHybrid{} is allowed to expand both. ``Gas Infrastructure'' combines gas facilities for power and heat production, ``\COtwo{} Infrastructure'' and ``\Htwo{} Infrastructure'' combine transport and storage for each carrier. ``Carbon Capture at Point Sources'' combines all technologies with integrated carbon capture, including the cost of the main facility (e.g., CHP unit) and the carbon capture application.}
    \label{fig:cost_bar}
\end{figure}

@@

System cost differences between the models are driven by wind and solar as well as carbon capture technologies. In scenarios with \carbon{} and \hydrogen{} transportation, the model primarily reduces the reliance on high-cost DAC from 392~Mt/a carbon removal in the \baselinescenario{} to 112~Mt/a in the \hybridscenario{} (see Fig.~\ref{fig:balance_captured_carbon}). At the same time, carbon capture from biomass combustion, gas-based industry processes and biogas-to-gas upgrading increase. The increased usage of upgraded biogas leads to a lower synthetic methane production.
% At the same time, the transport of \carbon{} and \hydrogen{} reduce the reliance on methanation and therefore the overall demand for CC volumes in the model. Another knock-on effect of less reliance on DAC is its reduced endogenous demand for dedicated renewables as well as heat pumps and biomass CHPs to supply power and heat.
\begin{figure}[ht!]
    \centering
    \includegraphics[width=\linewidth]{comparison/default/figures/90_nodes/balance_bar_carbon.png}
    \caption{Balance of captured carbon for all models in the Net-Zero scenario. Positive values indicate carbon capture, negative values indicate carbon consumption. By integrating \Htwo{} and \COtwo{} networks, the predominant method for carbon removal shifts from Direct Air Capture (DAC) to a bioenergetic process that incorporates carbon capture. At the same time the reliance on methanation decreases.}
    \label{fig:balance_captured_carbon}
\end{figure}

@@

In the \hydrogenscenario{}, the main hydrogen transportation routes go from Spain (50~GW) and the United Kingdom (18~GW) to Central Europe to supply local carbonaceous fuel production with captured carbon from point sources (see Fig~\ref{fig:balance_map}, bottom right). Also, the \Htwo{} network supplies spatially-fixed \Htwo{} demand for industry in Central Europe (2.5~GW in total) at much lower costs than local electrolysis or SMR. Carbon sequestration primarily locates in Portugal where large DAC facilities operate (14~kt/h, see Fig~\ref{fig:balance_map}, top right). \carbon{} prices are low in Central Europe (110-120~€/t) and highest at sequestration regions (160~€/t). Hydrogen prices have flattened out at a relatively low level of €60~/MWh in all regions.

@@

However, both transport systems provide individual cost-benefits to the system. While the carbon grid makes it possible to transport captured carbon for sequestration at low cost, the dependence on DAC near the coast causes higher costs in the hydrogen scenario{}. On the other hand, the \hydrogengrid{} is able to supply the spatially-fixed hydrogen demands in Central Europe with low-cost hydrogen imports, while the \modCO{} is forced to deploy electrolysis in regions with poor renewable resources as well as additional gas infrastructure in Central Europe. The latter aspect dominates the first, leading to lower system cost in the \hydrogenscenario{}.

@@

\begin{figure*}[ht!]
    \centering
    \begin{subfigure}{.5\textwidth}
        \centering
        \includegraphics[width=\linewidth]{full/figures/90_nodes/balance_map_carbon.png}
        \label{fig:balance_map_carbon_full}
    \end{subfigure}%
    \begin{subfigure}{.5\textwidth}
        \centering
        \includegraphics[width=\linewidth]{full/figures/90_nodes/balance_map_hydrogen.png}
        \label{fig:balance_map_hydrogen_full}
    \end{subfigure}
    \caption{Optimal operation, flows and prices of the carbon (left) and hydrogen (right) sectors for the \hybridmodel{} in the net zero scenario. For each region, upper semicircles show the average production per technology, lower semicircles the consumption, and colors the average marginal prices. Carbon Sequestration offshore are drawn as full circles. Lines and arrows show the interregional transportation. \carbon{} from point-source in the inland either supplies local CU with imported \hydrogen{} or facilitates sequestration in nearby offshore regions.
    % Carbon network looks the same as in~\cite{morbeeOptimisedDeploymentEuropean2012}: two backbones, one in the nothern Europe other in south east.
    }
    \label{fig:balance_map_full}
\end{figure*}

The \hybridscenario{}, the model combines the advantages of hydrogen and carbon networks, resulting in the highest system cost reduction of \label{}5.2\% (\label{}41~bn€/a) compared to the \modBase{}. As in the \hydrogenscenario{}, the model transports low-cost hydrogen to the point sources in Central Europe to supply CU, but also transports low-cost carbon from point sources close the shore to sequestration sites (see Fig.~\ref{fig:balance_map_full}). This reduces reliance on high-cost DAC and provides low-cost hydrogen across the regions.

For the \carbon{} network, the model employs short and direct pipeline routes to transport carbon from inland point sources to nearby sequestration sites, as for example in Northern Germany, Italy and Greece (see Fig.~\ref{fig:balance_map_full}, left). Additionally, the \COtwo{} network balances out small local variations in carbon demand and supply to produce CU, as for example in the middle of France.
The hydrogen routes are similar to those in the \hydrogenscenario{} with transports from Western regions (Iberian Peninsula and British Isles) to distributed CU production sites across the continent (see Fig.~\ref{fig:balance_map_full}, right).
While the topology of the \Htwo{} network (8~bn€ investments) roughly corresponds to that of the \modH{} primarily driven by CU, the \COtwo{} network plays more of a complementary role primarily driven by CS (3~bn€ investments, see Fig.~\ref{fig:cost_bar_transmission}).

\begin{figure}[ht]
    \centering
    \includegraphics[width=\linewidth]{comparison/default/figures/90_nodes/captureshare_line.png}
    \caption{Proportion of plants with integrated carbon capture for all models in the net zero scenario. The size of the dots corresponds to the average capacity factor of the respective technology. While peak load technologies such as gas-fired combined heat and power (CHP) plants, which are only in operation on a few days a year, are not expanded for carbon capture, baseload point sources such as biomass CHP and process emissions from industry are fully (partially) developed as soon as the transport of \COtwo{} (\Htwo{}) is permitted.}
    % A CO2 network unlock BECSS potentials, most biomass emissions are captured at point sources and transported to sequestration sites. cite~\cite{rosaAssessmentCarbonDioxide2021}
    \label{fig:captureshare_line}
\end{figure}%

In this combination, the networks enable all technologies that run on a high average, above 70\%, capacity factor to capture carbon (see Fig.~\ref{fig:captureshare_line}). Except for SMR and gas-fired Combined Heat and Power Plants (CHP) which operate a few weeks annually, all emissions from carbon-emitting technologies are captured.
% In comparison with the \hydrogenscenario, Biomass CHP supplies more flexible heat, power and relatively low-cost carbon.
Thus, only a few Direct Air Capture (DAC) installations are installed, such as those in southern Spain for methanolization plants. This leads to reduced power consumption and, therefore, less required wind and solar power deployment.

\begin{figure}[htb!]
    \centering
    \includegraphics[width=0.9\linewidth]{difference/comparison/emission-reduction-0.1/figures/90_nodes/cost_bar.png}
    \caption[short]{Net change in system cost when tightening the \COtwo{} emission target from net-zero to net-negative 10\% of 1990s emissions. Each cost bar is split into contribution of the same technology groups, except for \carbon{} sequestration cost, which are now displayed separately. For all models, Direct Air Capture (DAC) contributes most to the additional \carbon{} removal, requiring further solar, wind and heat pump capacities for electricity and heat input.}
    \label{fig:net-negative_cost_bar}
\end{figure}

>>>>

% make abbreviation for co2
\newcommand{\carbon}{CO$_2$}
\newcommand{\hydrogen}{H$_2$}
\newcommand{\carbongrid}{CO$_2$\=/Grid}
\newcommand{\hydrogengrid}{H$_2$\=/Grid}
\newcommand{\baselinescenario}{Baseline scenario}
\newcommand{\carbonscenario}{CO$_2$\=/Grid scenario}
\newcommand{\hydrogenscenario}{H$_2$\=/Grid scenario}
\newcommand{\hybridscenario}{Hybrid scenario}

@@

\title{Evaluating Hydrogen and Carbon Network Strategies for the European Energy System}
% \title{Competition and Synergies of \hydrogen{} and \carbon{} Networks in Europe}

@@

\begin{abstract}
    Hydrogen and carbon dioxide transport can both play an important role in climate-neutral energy systems. Hydrogen networks help serve regions with high energy demand, while excess emissions are transported away in carbon dioxide networks. Regarding the synthesis of carbonaceous fuels, it is less clear which input should be transported: hydrogen to carbon point sources or carbon to low-cost hydrogen. We explore the potential synergies and competition of both networks in a cost-optimal carbon-neutral European energy system. In direct comparison, a hydrogen network is more cost-effective than a carbon network, since it serves to transport hydrogen to demand and to point source of carbon for utilization. However, in a hybrid scenario where both networks are present, the carbon network effectively complements the hydrogen network, promoting carbon capture from biomass and reducing reliance on direct air capture. This study shows how the European energy system in future net-zero and net-negative carbon dioxide emissions scenarios benefits from integrating both networks.
\end{abstract}

@@

The transition to a climate-neutral European economy is a pressing challenge that demands coordinated action across various energy sectors. While management of both carbon dioxide (\carbon{}) and hydrogen (\hydrogen{}) is considered a critical component of this transition, a gap exists in understanding how new hydrogen infrastructure effectively interacts with carbon management technologies, including carbon capture, transport, storage, utilization, and sequestration. Hydrogen is considered in several sectors in industry not amenable to electrification, and offers an efficient way to transport and store energy over long distances and periods, and can be produced from electrolysis or steam methane reforming (SMR), for example. Captured carbon can be stored in geological formations to permanently extract it from the atmosphere, a process known as carbon sequestration (CS). Additionally, carbon can be combined with hydrogen to produce carbonaceous fuels to supply for example synthetic kerosene for aviation, synthetic methanol for shipping or synthetic methane for industrial feedstock. From hereon, we jointly refer to the production of these carbonaceous fuels as Carbon Utilization (CU). Carbon can be effectively captured from industrial processes and the combustion of biomass, fossil fuels, or synthetic carbonaceous fuels through point-source carbon capture (CC) techniques, or harvested from the atmosphere using direct air capture (DAC). Both \carbon{} and \hydrogen{} networks are likely to play a crucial role in the cost-effective integration of these technologies to enable net-zero economies.

Recently, policymakers and industry in Europe have started developing hydrogen and carbon management strategies~\cite{GermanyDevelopingStrategy2023,CarbonManagementStrategie}, planning infrastructure components~\cite{CONetz}, and committing to the first carbon utilization projects~\cite{EFuelsPilotPlant2022,OrstedAssumesFull,GROUNDBREAKINGEFUELPRODUCTION,DLREfuelsDLR}. With the European Union's goal of achieving climate neutrality by 2050, a wide range of programs, funding models and initiatives have been established with dedicated instruments supporting both the ramp-up of the hydrogen and carbon economy~\cite{eu2023netzero,europeangreendeal,europeaninnovationfund}.
Collaborative industry initiatives like the European Hydrogen Backbone~\cite{gasforclimateEuropeanHydrogenBackbone2022} or the Hydrogen Infrastructure Map~\cite{H2InfrastructureMap} showcase the potential of hydrogen as a fuel and energy carrier. Some natural gas pipelines have already been repurposed to transport hydrogen~\cite{RohrFreiFuer}. Business models from companies like Tree Energy Solutions~\cite{TESHydrogenLife2023}, Carbfix~\cite{WeTurnCO2}, and Equinor~\cite{adomaitisEquinorRWEBuild2023} advertise carbon management hubs that provide green hydrogen, methane, and synthetic fuels on the one hand and offer to purchase \carbon{} on the other hand. The Northern Lights project in Norway~\cite{NorthernLightsWhat} is planning with a transport and sequestration capacity of 1.5~Mt \carbon{} per year to be operative in 2024, expanding to a targeted scale of 5~Mt per year of sequestration by 2030.
The market potential for carbon capture and sequestration in Europe is underscored by the Capture Map's estimated potential~\cite{ToolsGreenTransition} of 1.7 Gt carbon annually from point sources, accounting for about half of the continent's emissions, as well as estimated sequestration capacity potentials of up to 3~Gt/a~\cite{europeancommissionEuropeanCO2Storage}. However, it's important to note that carbon sequestration technology, unlike carbon capture, is not yet fully mature and its full potential and implications remain to be explored in depth.
To advance the carbon economy, the Clean Air Task Force emphasizes the importance of developing a carbon transport system in Europe alongside a hydrogen network~\cite{lockwoodEuropeanStrategyCarbon}. Carbon pipelines, considered a mature technology, have seen widespread installations in the United States and Canada, primarily to supply \carbon{} to enhanced oil recovery~\cite{righettiSitingCarbonDioxide2017,friedmannNETZEROGEOSPHERICRETURN}.

Up to this point, it remains unclear how the two transport systems of hydrogen and carbon may complement or even compete with each other. Both can bridge the misalignment of sources and sinks for carbon and hydrogen. A hydrogen network can supply regions with geographically fixed hydrogen demand, such as for steel production, with hydrogen from regions with the best renewable resources as well as enable CU at the site of point-source CC. On the other hand, the carbon transport system allows for transporting captured carbon to regions with sequestration potentials or high-quality renewable resources, the latter enabling cost-effective CU.

In the literature, the two network approaches and underlying technologies have been discussed in several of publications, all of which, however, dealt with the isolated aspects~\cite{bakkenLinearModelsOptimization2008,morbeeOptimisedDeploymentEuropean2012,stewartFeasibilityEuropeanwideIntegrated2014,oeiModelingCarbonCapture2014,elahiMultiperiodLeastCost2014,burandtDecarbonizingChinaEnergy2019,middletonSimCCSOpensourceTool2020,bjerketvedtOptimalDesignCost2020,weiProposedGlobalLayout2021,damoreOptimalDesignEuropean2021,becattiniCarbonDioxideCapture2022,neumannBenefitsHydrogenNetwork2022}. Such techno-economic models, in contrast for example to Integrated Assessment Models, can account for the spatial distribution of carbon sources and sinks which are crucial to provide a holistic view of the energy system and its technological interactions. Neumann et al.~\cite{neumannBenefitsHydrogenNetwork2022} demonstrated the interchangeability of hydrogen and electricity grid expansion in a climate-neutral European energy system at the expense of higher investments. The underlying, highly-resolved model encompasses most relevant sectors, however it neglects the option to transport carbon across regions.
Morbee et al.~\cite{morbeeOptimisedDeploymentEuropean2012} optimizes the topology and capacity of a \carbon{} network in Europe, but only considers the power sector without co-optimizing renewable deployment. This limited sectoral scope cannot capture important dynamics of carbon management, since it neglects the sectors like industry that will need to handle most \carbon{} in the future.
A comprehensive example is found in~\cite{becattiniCarbonDioxideCapture2022}, which presents a mixed-integer model to optimize the time-evolution of a \carbon{} transport system in Switzerland, connecting to a remote sequestration site in Norway. However, this limited spatial scope fails to consider other sequestration sites and co-benefits from connecting the \carbon{} network to neighboring countries.
Other high-resolution energy system models do not feature detailed CU technologies and \carbon{} transport. Euro-Calliope, for example, does neither feature carbon nor hydrogen transport, and thus forces generic synthetic fuel production to rely on captive electrolysis and DAC in the same region~\cite{pickeringDiversityOptionsEliminate2022}.

No study has yet considered the co-optimization and comprehensive assessment of both \carbon{} and \hydrogen{} networks in a fully sector-coupled energy system. However, such an assessment is strongly needed to identify realms of competition and synergy between hydrogen and carbon management technologies. In this paper, we present a detailed study of the European energy system for 2050, which includes high geographical resolution and a comprehensive representation of carbon management technologies. The study is conducted using the PyPSA-Eur energy system model which optimizes operation and investments in all relevant energy sectors to supply projected energy demands for 2050 (see Fig.~\ref{fig:total-demand-bar}). It is assumed that Europe is self-sufficient in energy and does not import any fuels. This drives carbonaceous fuels production to be located at sites where both \carbon{} and \hydrogen{} can be provided at low cost.

@@

To investigate the competing transport dynamics between \carbon{} and \hydrogen{} networks, we contrast the following four model scenarios:
%
\begin{itemize}
    \item[] \textit{Baseline:} Neither \carbon{} nor \hydrogen{} networks are constructed.
    \item[] \textit{\carbongrid{}:} Only the \carbon{} network is developed.
    \item[] \textit{\hydrogengrid{}:} Only the \hydrogen{} network is developed.
    \item[] \textit{Hybrid:} Both \carbon{} and \hydrogen{} networks are developed.
\end{itemize}

@@

All scenarios share key energy system characteristics but differ in system costs and technology use (Fig.~\ref{fig:cost_bar}). Costs in the \baselinescenario{} are the highest at 779~bn€/a. Models with \carbon{} and \hydrogen{} transport options are more cost-effective, reducing costs by 3.1\% in the \carbongrid{}, 4.8\% in \hydrogengrid{}, and 5.2\% in \hybridscenario{}.

\begin{figure}[ht!]
    \centering
    \includegraphics[width=\linewidth]{comparison/default/figures/90_nodes/cost_bar.png}
    \caption[short]{Total annual system cost subdivided into groups of technologies for the different models of the European energy system with a net-zero emission target. While in the \baselinescenario{}, the model has neither a \carbon{} nor an \hydrogen{} network, it can expand both in the \hydrogenscenario{}. ``Gas Infrastructure'' combines gas facilities for power and heat production, ``\carbon{} Infrastructure'' and ``\hydrogen{} Infrastructure'' combine transport and storage for each carrier. ``Carbon Capture at Point Sources'' combines all technologies with integrated carbon capture, including the cost of the main facility (e.g., CHP unit) and the carbon capture application.}
    \label{fig:cost_bar}
\end{figure}

@@

System cost differences between the models are driven by wind and solar as well as carbon capture technologies. In scenarios with \carbon{} and \hydrogen{} transportation, the model primarily reduces the reliance on high-cost DAC from 392~Mt/a carbon removal in the \baselinescenario{} to 112~Mt/a in the \hybridscenario{} (see Fig.~\ref{fig:balance_captured_carbon}). At the same time, carbon capture from biomass combustion, gas-based industry processes and biogas-to-gas upgrading increase. The increased usage of upgraded biogas leads to a lower synthetic methane production.
% Another knock-on effect of less reliance on DAC is its reduced endogenous demand for dedicated renewables as well as heat pumps and biomass CHPs to supply power and heat.

\begin{figure}[ht!]
    \centering
    \includegraphics[width=\linewidth]{comparison/default/figures/90_nodes/balance_bar_carbon.png}
    \caption{Balance of captured carbon for all models in the Net-Zero scenario. Positive values indicate carbon capture, negative values indicate carbon consumption. By integrating \hydrogen{} and \carbon{} networks, the predominant method for carbon removal shifts from Direct Air Capture (DAC) to bio-energetic processes with capture. At the same time the reliance on methanation decreases.}
    \label{fig:balance_captured_carbon}
\end{figure}

@@

In the \hydrogenscenario{}, the main hydrogen transportation routes go from Spain (50~GW) and the United Kingdom (18~GW) to Central Europe to supply local carbonaceous fuel production with captured carbon from point sources (see Fig~\ref{fig:balance_map}, bottom right). Also, the \hydrogen{} network supplies spatially-fixed \hydrogen{} demand for industry in Central Europe (2.5~GW in total) at much lower costs than local electrolysis or SMR. Carbon sequestration primarily locates in Portugal where large DAC facilities operate (14~kt/h, see Fig~\ref{fig:balance_map}, top right). \carbon{} prices are low in Central Europe (110-120~€/t) and highest at sequestration regions (160~€/t). Hydrogen prices have flattened out at a relatively low level of €60~/MWh in all regions.

@@

However, both transport systems provide individual cost-benefits to the system. While the carbon grid makes it possible to transport captured carbon for sequestration at low cost, the dependence on DAC near the coast causes higher costs in the hydrogen scenario{}. On the other hand, the \hydrogengrid{} is able to supply the spatially-fixed hydrogen demands in Central Europe with low-cost hydrogen imports, while the model in the \carbonscenario{} is forced to deploy electrolysis in regions with poor renewable resources as well as additional gas infrastructure in Central Europe. The latter aspect dominates the first, leading to lower system cost in the \hydrogenscenario{}.

@@

\begin{figure*}[ht!]
    \centering
    \begin{subfigure}{.5\textwidth}
        \centering
        \includegraphics[width=\linewidth]{full/figures/90_nodes/balance_map_carbon.png}
        \label{fig:balance_map_carbon_full}
    \end{subfigure}%
    \begin{subfigure}{.5\textwidth}
        \centering
        \includegraphics[width=\linewidth]{full/figures/90_nodes/balance_map_hydrogen.png}
        \label{fig:balance_map_hydrogen_full}
    \end{subfigure}
    \caption{Optimal operation, flows and prices of the carbon (left) and hydrogen (right) sectors in the \hybridscenario{} assuming net \carbon{} neutrality. For each region, upper semicircles show the average production per technology, lower semicircles the consumption, and colors the average marginal prices. Carbon Sequestration offshore are drawn as full circles. Lines and arrows show the interregional transportation. \carbon{} from point-source in the inland either supplies local CU with imported \hydrogen{} or facilitates sequestration in nearby offshore regions.
    % Carbon network looks the same as in~\cite{morbeeOptimisedDeploymentEuropean2012}: two backbones, one in the nothern Europe other in south east.
    }
    \label{fig:balance_map_full}
\end{figure*}

The \hybridscenario{}, the model combines the advantages of hydrogen and carbon networks, resulting in the highest system cost reduction of 5.2\% (41~bn€/a) compared to the \baselinescenario{}. Again, the model transports low-cost hydrogen to the point sources in Central Europe to supply CU, but also transports low-cost carbon from point sources close the shore to sequestration sites (see Fig.~\ref{fig:balance_map_full}). This reduces reliance on high-cost DAC and provides low-cost hydrogen across the regions.

For the \carbon{} network, the model employs short and direct pipeline routes to transport carbon from inland point sources to nearby sequestration sites, as for example in Northern Germany, Italy and Greece (see Fig.~\ref{fig:balance_map_full}, left). Additionally, the \carbon{} network balances out small local variations in carbon demand and supply to produce CU, as for example in the middle of France.
The hydrogen routes are similar to those in the \hydrogenscenario{} with transports from Western regions (Iberian Peninsula and British Isles) to distributed CU production sites across the continent (see Fig.~\ref{fig:balance_map_full}, right).
While the topology of the \hydrogen{} network (8~bn€ investments) roughly corresponds to that of the \hydrogenscenario{} primarily driven by CU, the \carbon{} network plays more of a complementary role primarily driven by CS (3~bn€ investments, see Fig.~\ref{fig:cost_bar_transmission}).

\begin{figure}[ht]
    \centering
    \includegraphics[width=\linewidth]{comparison/default/figures/90_nodes/captureshare_line.png}
    \caption{Proportion of plants with integrated carbon capture for all models in the net zero scenario. The size of the dots corresponds to the average capacity factor of the respective technology. While peak load technologies such as gas-fired combined heat and power (CHP) plants, which are only in operation on a few days a year, are not expanded for carbon capture, baseload point sources such as biomass CHP and process emissions from industry are fully (partially) developed as soon as the transport of \carbon{} (\hydrogen{}) is permitted.}
    % A CO2 network unlock BECSS potentials, most biomass emissions are captured at point sources and transported to sequestration sites. cite~\cite{rosaAssessmentCarbonDioxide2021}
    \label{fig:captureshare_line}
\end{figure}%

In this combination, the networks enable all technologies that run on a high average, above 70\%, capacity factor to capture carbon (see Fig.~\ref{fig:captureshare_line}). Except for SMR and gas-fired Combined Heat and Power Plants (CHP) which operate a few weeks annually, all emissions from carbon-emitting technologies are captured.
Thus, only a few Direct Air Capture (DAC) installations are installed, such as those in southern Spain for methanolization plants. This leads to reduced power consumption and, therefore, less required wind and solar power deployment.

\begin{figure}[htb!]
    \centering
    \includegraphics[width=0.9\linewidth]{difference/comparison/emission-reduction-0.1/figures/90_nodes/cost_bar.png}
    \caption[short]{Net change in system cost when tightening the \carbon{} emission target from net-zero to net-negative 10\% of 1990s emissions. Each cost bar is split into contribution of the same technology groups, except for \carbon{} sequestration cost, which are now displayed separately. For all models, Direct Air Capture (DAC) contributes most to the additional \carbon{} removal, requiring further solar, wind and heat pump capacities for electricity and heat input.}
    \label{fig:net-negative_cost_bar}
\end{figure}
